---
title: "Final Project Guidelines"
author: "Data Sci 112"
embed-resources: true
---

```{r, include = FALSE}
set.seed(492)
```

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, message = FALSE, warning = FALSE)
```

# Overview

Your final project is your chance to demonstrate your knowledge from DS 112 in a real-world setting.  You will choose a dataset - or possibly multiple related datasets - and perform an analysis to answer research questions.

You will summarize your findings into a poster, and present it at one of two poster sessions:

* **Thursday, June 6th, 4-7pm**
* **Wednesday, June 12th, 8:30-11:30am**

Please note that if you choose the earlier time slot, you are *opting-in* to turning your project in almost a full week earlier than it would normally be due.  You also have Exam 2 on Wednesday, June 5th; so you'll need to manage your time carefully.

No extensions will be granted for projects, except in *highly* unusual extenuating circumstances.  Please start early!

Projects may be done in groups of **1,2, or 3** students.  The guidelines are the same regardless; that is, if you choose to work alone, I will expect the same level of quality as if you choose to work in a group of 3.

# Checkpoint 1: Project Proposal

To make sure you are on track, you have a two-page project proposal due on **Friday, May 17th**.

This proposal must be two pages only, and it must contain:

1. Your group member names.

2. Information about the dataset(s) you intend to analyze: 
    + Where is the data located?
    + Who collected the data and why?
    + What information (variables) are in the dataset?
    
3. Research Questions: You should have one **primary** research question and a few **secondary** ones

4. Preliminary exploration of your datasets:  A few simple **plots** or **summary statistics** that relate to the variables you will study.


### Finding Data

Some good places to find datasets are:

* [Kaggle](www.kaggle.com)

* [Tidy Tuesday](https://github.com/rfordatascience/tidytuesday)

* [Google Data Set Search](https://toolbox.google.com/datasetsearch)

* [Reddit Datasets](https://www.reddit.com/r/datasets/)

* [List of JSON APIs](https://github.com/toddmotto/public-apis)

* [Project Gutenberg (good source of textual data)](https://www.gutenberg.org/)

* State or federal government websites, such as the [U.S. Government's Open Data](https://www.data.gov/)

# Final Poster

[Here](https://slidesgo.com/research-poster?format=4%3A3) are some templates that you can (but are not required to) use. Your poster will need to be printed on 24" x 36" paper. You may select all the most basic printing options (e.g., matte paper, no lamination, not mounted); or, you may upload your poster the day before your presentation to have us print it for you.

Your posters will be graded on:

* **Presentation:**
    + Was the poster clear and visually appealing?
    + Do the visualizations tell a story?
    + Were the presenters able to describe the project and answer questions.

* **Correctness:**
    + Were the correct summaries and analyses chosen for the data?
    + Were the results correctly interpreted and described?
    + Was the choice of analysis reasonable for the research question?

* **Complexity:**
    + Did the project address multiple rich aspects of the dataset(s)?
    + Did the project use skills from across the quarter to answer questions?
    + Did the project incorporate something from Week 9 or 10?  (Webscraping, APIs, data joining, chloropleths...) or from beyond the course?

* **Data insight:**
    + Were the research questions interesting and insightful?
    + Did the project make clear and relevant data conclusions?
    + Does the project tell a story with real-world impact?